Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-17 paper/paper.tex
shrink related work even more

Tesson et al.\ demonstrated the efficacy of leveraging Coq to establish an
approach for implementing a robust system dedicated to verifying the correctness
of program transformations for functions that manipulate
lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does not
include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted OCaml code
from their formalisation. Larchey-Wendling and Monin encode recursion schemes in
Coq, by formalising computational graphs of algorithms~\cite{larchey2022braga}.
Their work does not focus on encoding generic recursion schemes, and proving
their algebraic laws. Castro-Perez et al.~\cite{farmsCastro} encode the laws of
hylomorphisms as part of the type system of a functional language to calculate
parallel programs from specifications. Their work focuses on parallelism, and
they do not formalise their approach in a proof assistant, and the laws of
hylomorphisms are axioms in their system.

@@

Fusion in functional programming is a general technique for combining multiple
computations into a single pass to eliminate intermediate data structures or
repeated traversals, improving efficiency. The fusion we have discussed in this
paper relating to program calculation is one form of fusion, but there are other
approaches, and prior research has explored the mechanisation and verification
of fusion transformations.

Fusion has been applied to recursive tree traversals, in order to reduce the
number of times each section of the tree must be visited. Research has
demonstrated that this sort of fusion can be performed automatically by a
compiler~\cite{10.1145/3133900}, and the correctness of tree traversal fusion
has been verified mechanically in Coq~\cite{10.1145/3377555.3377884}.

Another related family transformations are worker/wrapper
transformations~\cite{gill09}, which have been used in the context of optimising
compilers, and have been mechanised in Agda~\cite{SCULTHORPE_HUTTON_2014}. This
technique has been applied to improving the performance of programs defined
using folds via fusion~\cite{hutton10}. Their work does not focus on the use of
an interactive proof assistant, but rather on the manual use of equational
reasoning to re-write programs and on the automatic transformation of functional
programs by a compiler.

@@

%In particular,
The problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
Take, for instance, division by iterated subtraction, which is inherently
non-structurally recursive since it involves recursion based on the result of a
subtraction. There is a nonstandard implementation of divivion found on Coq's
standard library, which involves a four-argument function that effectively
combines subtraction and division. Similarly, the mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognizable as a form of
mergesort''~\cite{AbreuDHJMS23}.
There are common approaches in Coq to deal with termination, but none of these
approaches address program calculation techniques, and the mechanisation of
fusion laws.

% \subsection{Fusion}\label{sec:fusion}

% Fusion in functional programming is a general technique for combining multiple
% computations into a single pass to eliminate intermediate data structures and
% improve efficiency. The fusion we have discussed in this paper relating to
% program calculation is one form of fusion, but there are other approaches.

% \emph{Deforestation}, introduced by Wadler~\cite{wadler88}, is a specific
% approach to fusion that aims to remove intermediate trees (such as lists)
% created between chained operations. A well-known method of deforestation is
% \haskell{foldr/build} fusion~\cite{gill93}, which optimizes list-processing by
% fusing list-producing functions (like \haskell{map} or \haskell{filter}) with
% list-consuming ones (like \haskell{foldr}), avoiding the creation of temporary
% lists.
% %
% Extending this idea, stream fusion generalizes the concept to handle other data
% types, using a lazy stream abstraction to represent sequences and fuse
% transformations like \haskell{map} and \haskell{fold}~\cite{coutts07}.
% Both approaches eliminate redundant structures and ensure computations happen in
% a single efficient pass over the data.

% Another dimension of fusion is \emph{tupling}, a technique used to combine
% multiple recursive functions into a single function by grouping their results
% into a tuple~\cite{hughes89}. This approach helps avoid redundant traversals of
% the same data structure when multiple operations are applied.
% In addition to
% techniques like \haskell{foldr/build} fusion and stream fusion, tupling
% complements deforestation by allowing multiple recursive functions to share a
% single traversal, further optimizing computations and minimizing intermediate
% data structures.

% In the context of Coq, Danvy explains the use of fold/unfold lemmas to mechanize
% equational reasoning about recursive programs~\cite{DANVY_2022}. Fold-unfold
% lemmas enable optimizations like fusion by allowing programs to be
% systematically simplified, including eliminating intermediate structures.
% Additionally, an approach to fusion called the worker/wrapper
% transformation~\cite{gill09} was formalised and mechanised in
% Agda~\cite{SCULTHORPE_HUTTON_2014}.

>>>>

% Tesson et al.\ demonstrated the efficacy of leveraging Coq to establish an
% approach for implementing a robust system dedicated to verifying the correctness
% of program transformations for functions that manipulate
% lists~\cite{10.1007/978-3-642-17796-5_10}.
Tesson et al.\ used Coq to mechanically verify program transformations for functions that manipulate lists~\cite{10.1007/978-3-642-17796-5_10}.
Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does not
include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted OCaml code
from their formalisation. Larchey-Wendling and Monin encode recursion schemes in
Coq, by formalising computational graphs of algorithms~\cite{larchey2022braga}.
Their work does not focus on encoding generic recursion schemes, and proving
their algebraic laws. Castro-Perez et al.~\cite{farmsCastro} encode the laws of
hylomorphisms as part of the type system of a functional language to calculate
parallel programs from specifications. Their work focuses on parallelism, and
they do not formalise their approach in a proof assistant, and the laws of
hylomorphisms are axioms in their system.

@@

% Fusion in functional programming is a general technique for combining multiple
% computations into a single pass to eliminate intermediate data structures or
% repeated traversals, improving efficiency.
The fusion we have discussed in this
paper relating to program calculation is one form of fusion, but there are other
approaches.
%
Fusion has been applied to recursive tree traversals, in order to reduce the
number of times each section of the tree must be visited. Research has
demonstrated that this sort of fusion can be performed automatically by a
compiler~\cite{10.1145/3133900}, and the correctness of tree traversal fusion
has been verified mechanically in Coq~\cite{10.1145/3377555.3377884}.
%
Another related family transformations are worker/wrapper
transformations~\cite{gill09}, which have been used in the context of optimising
compilers, and have been mechanised in Agda~\cite{SCULTHORPE_HUTTON_2014}. This
technique has been applied to improving the performance of programs defined
using folds via fusion~\cite{hutton10}. Their work does not focus on the use of
an interactive proof assistant, but rather on the automatic optimisation of
programs by a compiler.

@@

%In particular,
The problem of \emph{nonstructural recursion} (including
divide-and-conquer algorithms) is well-studied~\cite{BoveKS16}.
% A distinction between standard and
% nonstandard recursion lies in the fact that standard recursion adheres strictly
% to recursing on pattern variables, avoiding reliance on results from other
% function calls. Nonstandard recursion may deviate from this convention, allowing
% for more flexibility.
Certain functions that are not structurally recursive can be reformulated using
a nonstandard approach to achieve structural recursion~\cite{AbreuDHJMS23}.
% Take, for instance, division by iterated subtraction, which is inherently
% non-structurally recursive since it involves recursion based on the result of a
% subtraction. There is a nonstandard implementation of divivion found on Coq's
% standard library, which involves a four-argument function that effectively
% combines subtraction and division.
% Similarly,
The mergesort in Coq's standard
library uses an ``explicit stack of pending merges'' in order to avoid issues
with nonstructural definitions. There is a major downside, however; as noted by
Abreu et al., the result is ``barely recognizable as a form of
mergesort''~\cite{AbreuDHJMS23}.
There are common approaches in Coq to deal with termination, but none of these
approaches address program calculation techniques, and the mechanisation of
fusion laws.
